Component response rate variation underlies the stability of complex systems

### 2020-03-13 notebook/ms.Rmd
Edit the introduction to smooth the writing and fix the notation again, and add a bit to Figure 1 legend

---
title: "Component response rate variation underlies the stability of complex systems"
author: "A. Bradley Duthie    ( alexander.duthie@stir.ac.uk )"
date: Biological and Environmental Sciences, University of Stirling, Stirling, UK,
  FK9 4LA
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document: default
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - nature.csl
    reference_docx: docx_template.docx
bibliography: references.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage[utf8]{inputenc}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 2
link-citations: yes
linkcolor: blue
csl: nature.csl
biblio-style: apalike
---

@@

The stability of a complex system generally decreases with increasing system size and interconnectivity, a counterintuitive result of widespread importance across the physical, life, and social sciences. Despite recent interest in the relationship between system properties and stability, the effect of variation in the response rate of individual system components remains unconsidered. Here I vary the component response rates ($\boldsymbol{\gamma}$) of randomly generated complex systems. I use numerical simulations to show that when component response rates vary, the potential for system stability is markedly increased. These results are robust to common network structures, including small-world and scale-free networks, and cascade food webs. Variation in $\boldsymbol{\gamma}$ is especially important for stability in highly complex systems, in which the probability of stability would otherwise be negligible. At such extremes of simulated system complexity, the largest stable complex systems would be unstable if not for $\boldsymbol{Var(\gamma)}$. My results therefore reveal a previously unconsidered aspect of system stability that is likely to be pervasive across all realistic complex systems.

@@

Randomly assembled complex systems can be represented as large square matrices ($\mathbf{M}$) with $S$ components (e.g., networks of species [@Allesina2012] or banks [@Haldane2011]). One element of such a matrix, $M_{ij}$, defines how component $j$ affects component $i$ in the system at a point of equilibrium [@Allesina2012]. Off-diagonal elements ($i \neq j$) therefore define interactions between components, while diagonal elements ($i = j$) define component self-regulation ($d$, e.g., carrying capacity in ecological communities). Traditionally, off-diagonal elements are assigned non-zero values with a probability $C$, which are sampled from a distribution with variance $\sigma_{M}^{2}$; diagonal elements are set to $d = -1$ [@May1972; @Allesina2012; @Allesina2015]. Local system stability is assessed using eigenanalysis, with the system being stable if the real parts of all eigenvalues ($\lambda$) of $\mathbf{M}$ are negative ($\Re(\lambda_{max}) < 0$) [@May1972; @Allesina2012]. In a large system (high $S$), eigenvalues are distributed uniformly [@Tao2010] within a circle centred at $\Re = -1$ (the mean value of diagonal elements) and $\Im = 0$, with a radius of $\sigma_{M}\sqrt{SC}$ [@May1972; @Allesina2012; @Allesina2015] (Figs 1a and 2a). Local stability of randomly assembled systems therefore becomes increasingly unlikely as $S$, $C$, and $\sigma_{M}$ increase.

May's [@May1972; @Allesina2012] stability criterion $\sigma_{M}\sqrt{SC} < 1$ assumes that the expected response rates ($\gamma$) of individual components to perturbations of the system are identical, but this is highly unlikely in any complex system. In ecological communities, for example, the rate at which population density changes following perturbation will depend on the generation time of organisms, which might vary by orders of magnitude among species. Species with short generation times will respond quickly (high $\gamma$) to perturbations relative to species with long generation times (low $\gamma$). Similarly, the speed at which individual banks respond to perturbations in financial networks, or individuals or institutions respond to perturbations in complex social networks, is likely to vary. The effect of such variance on stability has not been investigated in complex systems theory. Intuitively, variation in $\gamma$ ($\sigma^{2}_{\gamma}$) might be expected to decrease system stability by introducing a new source of variation into the system and thereby increasing $\sigma_{M}$. Here I show that, despite higher $\sigma_{M}$, realistic complex systems (such that $S$ is high but finite) are actually more likely to be stable if their individual component response rates vary. My results are robust across commonly observed network structures, including random [@May1972], small-world [@Watts1998], scale-free [@Albert2002], cascade food web [@Solow1998; @Williams2000] networks.

@@

**Figure 1: Eigenvalue distributions of random complex systems.** Each panel shows the real (x-axis) and imaginary (y-axis) parts of $S =$ `r S` eigenvalues from random $S \times S$ matrices. ($\textbf{a}$) A system represented by a matrix $\mathbf{A}$, in which all elements are sampled from a normal distribution with $\mu = 0$ and $\sigma_{A} = 1/\sqrt{S}$. Points are uniformly distributed within a circle centred at the origin with a radius of $\sigma_{A} \sqrt{S} = 1$. ($\textbf{b}$) The same system after including variation in the response rates of $S$ components, represented by the diagonal matrix $\gamma$, such that $\mathbf{M} = \gamma\mathbf{A}$. Elements of $\gamma$ are randomly sampled from a uniform distribution from $\min = 0$ to $\max = 2$. Eigenvalues of $\mathbf{M}$ are then distributed non-uniformly within a circle centred at the origin with a radius of $\sqrt{\sigma^{2}_{A}(1 + \sigma^{2}_{\gamma})S} \approx$ `r round(sqrt(joint_var) * sqrt(S), digits = 2);`. ($\textbf{c}$) A different random system $\mathbf{A}$ constructed from the same parameters as in $\textbf{a}$, except with diagonal values of $-1$. ($\textbf{d}$) The same system $\textbf{c}$ after including variation in component response rates, sampled from $\mathcal{U}(0, 2)$ as in $\textbf{b}$.

>>>>

---
title: "Component response rate variation underlies the stability of complex systems"
author: "A. Bradley Duthie    ( alexander.duthie@stir.ac.uk )"
date: Biological and Environmental Sciences, University of Stirling, Stirling, UK,
  FK9 4LA
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document: default
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - nature.csl
    reference_docx: docx_template.docx
bibliography: references.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage[utf8]{inputenc}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 1
link-citations: yes
linkcolor: blue
csl: nature.csl
biblio-style: apalike
---

@@

The stability of a complex system generally decreases with increasing system size and interconnectivity, a counterintuitive result of widespread importance across the physical, life, and social sciences. Despite recent interest in the relationship between system properties and stability, the effect of variation in response rate across system components remains unconsidered. Here I vary the component response rates ($\boldsymbol{\gamma}$) of randomly generated complex systems. I use numerical simulations to show that when component response rates vary, the potential for system stability increases. These results are robust to common network structures, including small-world and scale-free networks, and cascade food webs. Variation in $\boldsymbol{\gamma}$ is especially important for stability in highly complex systems, in which the probability of stability would otherwise be negligible. At such extremes of simulated system complexity, the largest stable complex systems would be unstable if not for variation in $\boldsymbol{\gamma}$. My results therefore reveal a previously unconsidered aspect of system stability that is likely to be pervasive across all realistic complex systems.

@@

Randomly assembled complex systems can be represented as large square matrices ($\mathbf{M}$) with $S$ components (e.g., networks of species [@Allesina2012] or banks [@Haldane2011]). One element of such a matrix, $M_{ij}$, defines how component $j$ affects component $i$ in the system at a point of equilibrium [@Allesina2012]. Off-diagonal elements ($i \neq j$) therefore define interactions between components, while diagonal elements ($i = j$) define component self-regulation (e.g., carrying capacity in ecological communities). Traditionally, off-diagonal elements are assigned non-zero values with a probability $C$, which are sampled from a distribution with variance $\sigma^{2}$; diagonal elements are set to $-1$ [@May1972; @Allesina2012; @Allesina2015]. Local system stability is assessed using eigenanalysis on $\mathbf{M}$, with the system being stable if the real parts of all eigenvalues ($\lambda$), and therefore the leading eigenvalue ($\lambda_{max}$), are negative ($\Re(\lambda_{max}) < 0$) [@May1972; @Allesina2012]. In a large system (high $S$), eigenvalues are distributed uniformly [@Tao2010] within a circle centred at $\Re = -d$ ($-d$ is the mean value of diagonal elements) and $\Im = 0$, with a radius of $\sigma\sqrt{SC}$ [@May1972; @Allesina2012; @Allesina2015] (Figs 1a and 2a). Local stability of randomly assembled systems therefore becomes increasingly unlikely as $S$, $C$, and $\sigma$ increase.

May's [@May1972; @Allesina2012] stability criterion $\sigma\sqrt{SC} < d$ assumes that the expected response rates ($\gamma$) of individual components to perturbations of the system are identical, but this is highly unlikely in any complex system. In ecological communities, for example, the rate at which population density changes following perturbation will depend on the generation time of organisms, which might vary by orders of magnitude among species. Species with short generation times will respond quickly (high $\gamma$) to perturbations relative to species with long generation times (low $\gamma$). Similarly, the speed at which individual banks respond to perturbations in financial networks, or individuals or institutions respond to perturbations in complex social networks, is likely to vary. The effect of such variance on stability has not been investigated in complex systems theory. Intuitively, variation in $\gamma$ ($\sigma^{2}_{\gamma}$) might be expected to decrease system stability by introducing a new source of variation into the system and thereby increasing $\sigma$. Here I show that, despite higher $\sigma$, realistic complex systems (in which $S$ is high but finite) are actually more likely to be stable if their individual component response rates vary. My results are robust across commonly observed network structures, including random [@May1972], small-world [@Watts1998], scale-free [@Albert2002], cascade food web [@Solow1998; @Williams2000] networks.

@@

**Figure 1: Eigenvalue distributions of random complex systems.** Each panel shows the real (x-axis) and imaginary (y-axis) parts of $S =$ `r S` eigenvalues from random $S \times S$ matrices. ($\textbf{a}$) A system represented by a matrix $\mathbf{A}$, in which all elements are sampled from a normal distribution with $\mu = 0$ and $\sigma_{A} = 1/\sqrt{S}$. Points are uniformly distributed within a circle centred at the origin with a radius of $\sigma_{A} \sqrt{S} = 1$. ($\textbf{b}$) The same system as $\textbf{a}$ after including variation in the response rates of $S$ components, represented by the diagonal matrix $\gamma$, such that $\mathbf{M} = \gamma\mathbf{A}$. Elements of $\gamma$ are randomly sampled from a uniform distribution from $\min = 0$ to $\max = 2$. Eigenvalues of $\mathbf{M}$ are then distributed non-uniformly within a circle centred at the origin with a radius of $\sqrt{\sigma^{2}_{A}(1 + \sigma^{2}_{\gamma})S} \approx$ `r round(sqrt(joint_var) * sqrt(S), digits = 2);`. ($\textbf{c}$) A different random system $\mathbf{A}$ constructed from the same parameters as in $\textbf{a}$, except with diagonal values of $-1$. ($\textbf{d}$) The same system $\textbf{c}$ after including variation in component response rates, sampled from $\mathcal{U}(0, 2)$ as in $\textbf{b}$.

### 2020-03-13 notebook/ms.tex
Edit the introduction to smooth the writing and fix the notation again, and add a bit to Figure 1 legend

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{setspace}
\setstretch{2}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            pdftitle={Component response rate variation underlies the stability of complex systems},
            pdfauthor={A. Bradley Duthie ( alexander.duthie@stir.ac.uk )},
            colorlinks=true,
            linkcolor=blue,
            citecolor=Blue,
            urlcolor=Blue,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

The stability of a complex system generally decreases with increasing
system size and interconnectivity, a counterintuitive result of
widespread importance across the physical, life, and social sciences.
Despite recent interest in the relationship between system properties
and stability, the effect of variation in the response rate of
individual system components remains unconsidered. Here I vary the
component response rates (\(\boldsymbol{\gamma}\)) of randomly generated
complex systems. I use numerical simulations to show that when component
response rates vary, the potential for system stability is markedly
increased. These results are robust to common network structures,
including small-world and scale-free networks, and cascade food webs.
Variation in \(\boldsymbol{\gamma}\) is especially important for
stability in highly complex systems, in which the probability of
stability would otherwise be negligible. At such extremes of simulated
system complexity, the largest stable complex systems would be unstable
if not for \(\boldsymbol{Var(\gamma)}\). My results therefore reveal a
previously unconsidered aspect of system stability that is likely to be
pervasive across all realistic complex systems.

@@

Randomly assembled complex systems can be represented as large square
matrices (\(\mathbf{M}\)) with \(S\) components (e.g., networks of
species\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}} or
banks\textsuperscript{\protect\hyperlink{ref-Haldane2011}{11}}). One
element of such a matrix, \(M_{ij}\), defines how component \(j\)
affects component \(i\) in the system at a point of
equilibrium\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}.
Off-diagonal elements (\(i \neq j\)) therefore define interactions
between components, while diagonal elements (\(i = j\)) define component
self-regulation (\(d\), e.g., carrying capacity in ecological
communities). Traditionally, off-diagonal elements are assigned non-zero
values with a probability \(C\), which are sampled from a distribution
with variance \(\sigma_{M}^{2}\); diagonal elements are set to
\(d = -1\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}}.
Local system stability is assessed using eigenanalysis, with the system
being stable if the real parts of all eigenvalues (\(\lambda\)) of
\(\mathbf{M}\) are negative
(\(\Re(\lambda_{max}) < 0\))\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}.
In a large system (high \(S\)), eigenvalues are distributed
uniformly\textsuperscript{\protect\hyperlink{ref-Tao2010}{14}} within a
circle centred at \(\Re = -1\) (the mean value of diagonal elements) and
\(\Im = 0\), with a radius of
\(\sigma_{M}\sqrt{SC}\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}}
(Figs 1a and 2a). Local stability of randomly assembled systems
therefore becomes increasingly unlikely as \(S\), \(C\), and
\(\sigma_{M}\) increase.

May's\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}
stability criterion \(\sigma_{M}\sqrt{SC} < 1\) assumes that the
expected response rates (\(\gamma\)) of individual components to
perturbations of the system are identical, but this is highly unlikely
in any complex system. In ecological communities, for example, the rate
at which population density changes following perturbation will depend
on the generation time of organisms, which might vary by orders of
magnitude among species. Species with short generation times will
respond quickly (high \(\gamma\)) to perturbations relative to species
with long generation times (low \(\gamma\)). Similarly, the speed at
which individual banks respond to perturbations in financial networks,
or individuals or institutions respond to perturbations in complex
social networks, is likely to vary. The effect of such variance on
stability has not been investigated in complex systems theory.
Intuitively, variation in \(\gamma\) (\(\sigma^{2}_{\gamma}\)) might be
expected to decrease system stability by introducing a new source of
variation into the system and thereby increasing \(\sigma_{M}\). Here I
show that, despite higher \(\sigma_{M}\), realistic complex systems
(such that \(S\) is high but finite) are actually more likely to be
stable if their individual component response rates vary. My results are
robust across commonly observed network structures, including
random\textsuperscript{\protect\hyperlink{ref-May1972}{1}},
small-world\textsuperscript{\protect\hyperlink{ref-Watts1998}{15}},
scale-free\textsuperscript{\protect\hyperlink{ref-Albert2002}{16}},
cascade food
web\textsuperscript{\protect\hyperlink{ref-Solow1998}{17},\protect\hyperlink{ref-Williams2000}{18}}
networks.

@@

\textbf{Figure 1: Eigenvalue distributions of random complex systems.}
Each panel shows the real (x-axis) and imaginary (y-axis) parts of
\(S =\) 400 eigenvalues from random \(S \times S\) matrices.
(\(\textbf{a}\)) A system represented by a matrix \(\mathbf{A}\), in
which all elements are sampled from a normal distribution with
\(\mu = 0\) and \(\sigma_{A} = 1/\sqrt{S}\). Points are uniformly
distributed within a circle centred at the origin with a radius of
\(\sigma_{A} \sqrt{S} = 1\). (\(\textbf{b}\)) The same system after
including variation in the response rates of \(S\) components,
represented by the diagonal matrix \(\gamma\), such that
\(\mathbf{M} = \gamma\mathbf{A}\). Elements of \(\gamma\) are randomly
sampled from a uniform distribution from \(\min = 0\) to \(\max = 2\).
Eigenvalues of \(\mathbf{M}\) are then distributed non-uniformly within
a circle centred at the origin with a radius of
\(\sqrt{\sigma^{2}_{A}(1 + \sigma^{2}_{\gamma})S} \approx\) 1.18.
(\(\textbf{c}\)) A different random system \(\mathbf{A}\) constructed
from the same parameters as in \(\textbf{a}\), except with diagonal
values of \(-1\). (\(\textbf{d}\)) The same system \(\textbf{c}\) after
including variation in component response rates, sampled from
\(\mathcal{U}(0, 2)\) as in \(\textbf{b}\).

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{setspace}
\setstretch{1}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            pdftitle={Component response rate variation underlies the stability of complex systems},
            pdfauthor={A. Bradley Duthie ( alexander.duthie@stir.ac.uk )},
            colorlinks=true,
            linkcolor=blue,
            citecolor=Blue,
            urlcolor=Blue,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

The stability of a complex system generally decreases with increasing
system size and interconnectivity, a counterintuitive result of
widespread importance across the physical, life, and social sciences.
Despite recent interest in the relationship between system properties
and stability, the effect of variation in response rate across system
components remains unconsidered. Here I vary the component response
rates (\(\boldsymbol{\gamma}\)) of randomly generated complex systems. I
use numerical simulations to show that when component response rates
vary, the potential for system stability increases. These results are
robust to common network structures, including small-world and
scale-free networks, and cascade food webs. Variation in
\(\boldsymbol{\gamma}\) is especially important for stability in highly
complex systems, in which the probability of stability would otherwise
be negligible. At such extremes of simulated system complexity, the
largest stable complex systems would be unstable if not for variation in
\(\boldsymbol{\gamma}\). My results therefore reveal a previously
unconsidered aspect of system stability that is likely to be pervasive
across all realistic complex systems.

@@

Randomly assembled complex systems can be represented as large square
matrices (\(\mathbf{M}\)) with \(S\) components (e.g., networks of
species\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}} or
banks\textsuperscript{\protect\hyperlink{ref-Haldane2011}{11}}). One
element of such a matrix, \(M_{ij}\), defines how component \(j\)
affects component \(i\) in the system at a point of
equilibrium\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}.
Off-diagonal elements (\(i \neq j\)) therefore define interactions
between components, while diagonal elements (\(i = j\)) define component
self-regulation (e.g., carrying capacity in ecological communities).
Traditionally, off-diagonal elements are assigned non-zero values with a
probability \(C\), which are sampled from a distribution with variance
\(\sigma^{2}\); diagonal elements are set to
\(-1\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}}.
Local system stability is assessed using eigenanalysis on
\(\mathbf{M}\), with the system being stable if the real parts of all
eigenvalues (\(\lambda\)), and therefore the leading eigenvalue
(\(\lambda_{max}\)), are negative
(\(\Re(\lambda_{max}) < 0\))\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}.
In a large system (high \(S\)), eigenvalues are distributed
uniformly\textsuperscript{\protect\hyperlink{ref-Tao2010}{14}} within a
circle centred at \(\Re = -d\) (\(-d\) is the mean value of diagonal
elements) and \(\Im = 0\), with a radius of
\(\sigma\sqrt{SC}\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}}
(Figs 1a and 2a). Local stability of randomly assembled systems
therefore becomes increasingly unlikely as \(S\), \(C\), and \(\sigma\)
increase.

May's\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}
stability criterion \(\sigma\sqrt{SC} < d\) assumes that the expected
response rates (\(\gamma\)) of individual components to perturbations of
the system are identical, but this is highly unlikely in any complex
system. In ecological communities, for example, the rate at which
population density changes following perturbation will depend on the
generation time of organisms, which might vary by orders of magnitude
among species. Species with short generation times will respond quickly
(high \(\gamma\)) to perturbations relative to species with long
generation times (low \(\gamma\)). Similarly, the speed at which
individual banks respond to perturbations in financial networks, or
individuals or institutions respond to perturbations in complex social
networks, is likely to vary. The effect of such variance on stability
has not been investigated in complex systems theory. Intuitively,
variation in \(\gamma\) (\(\sigma^{2}_{\gamma}\)) might be expected to
decrease system stability by introducing a new source of variation into
the system and thereby increasing \(\sigma\). Here I show that, despite
higher \(\sigma\), realistic complex systems (in which \(S\) is high but
finite) are actually more likely to be stable if their individual
component response rates vary. My results are robust across commonly
observed network structures, including
random\textsuperscript{\protect\hyperlink{ref-May1972}{1}},
small-world\textsuperscript{\protect\hyperlink{ref-Watts1998}{15}},
scale-free\textsuperscript{\protect\hyperlink{ref-Albert2002}{16}},
cascade food
web\textsuperscript{\protect\hyperlink{ref-Solow1998}{17},\protect\hyperlink{ref-Williams2000}{18}}
networks.

@@

\textbf{Figure 1: Eigenvalue distributions of random complex systems.}
Each panel shows the real (x-axis) and imaginary (y-axis) parts of
\(S =\) 400 eigenvalues from random \(S \times S\) matrices.
(\(\textbf{a}\)) A system represented by a matrix \(\mathbf{A}\), in
which all elements are sampled from a normal distribution with
\(\mu = 0\) and \(\sigma_{A} = 1/\sqrt{S}\). Points are uniformly
distributed within a circle centred at the origin with a radius of
\(\sigma_{A} \sqrt{S} = 1\). (\(\textbf{b}\)) The same system as
\(\textbf{a}\) after including variation in the response rates of \(S\)
components, represented by the diagonal matrix \(\gamma\), such that
\(\mathbf{M} = \gamma\mathbf{A}\). Elements of \(\gamma\) are randomly
sampled from a uniform distribution from \(\min = 0\) to \(\max = 2\).
Eigenvalues of \(\mathbf{M}\) are then distributed non-uniformly within
a circle centred at the origin with a radius of
\(\sqrt{\sigma^{2}_{A}(1 + \sigma^{2}_{\gamma})S} \approx\) 1.21.
(\(\textbf{c}\)) A different random system \(\mathbf{A}\) constructed
from the same parameters as in \(\textbf{a}\), except with diagonal
values of \(-1\). (\(\textbf{d}\)) The same system \(\textbf{c}\) after
including variation in component response rates, sampled from
\(\mathcal{U}(0, 2)\) as in \(\textbf{b}\).
